Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-10-23 documents/mixtures/mixture_models.tex
sersic index discussion; typos

% to-do
% -----
% - HOGG: go through remining github issues and close or fix
% - HOGG: send to Rix and Peng and Lupton and Bundy for comments
% - HOGG: submit!
% - HOGG: can we drop the regularization in the badness and still succeeed?  If so, let's do that (and drop it in the text).
% - HOGG: make what I say about M=6 lux and M=8 lux true -- ie, edit the Tractor code!
% - DSTN: can we be more specific about how Bosch is using them? I would like to cite users.
% - HOGG: should we make Machine Readable Tables for PASP?  Yes.

@@

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian (MoG)
modeling.  We have pioneered this in the area of distribution modeling
in large numbers of dimensions (\citealt{xd, xdqso, xdqsoz}), where
convolution occurs because the true or noise-free distribution is
convolved with the noise before being observed.  Here we are going to
capitalize on the convolution properties of MoGs in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is always
easier with Gaussians than cuspy profiles, but is \emph{particularly}
useful in data sets in which the PSF itself has also been modeled as a
MoG, which is not uncommon (for example, the \project{Sloan Digital
  Sky Survey} imaging pipelines make approximate MoG PSF models for
every imaging field; \citealt{lupton}).

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\project{Sloan Digital Sky Survey} (\sdss) Catalog galaxy models,
which only include exponential, de~Vaucouleurs, and composite (mixture
of the two) radial profiles.  In detail, in fact, the \sdss\ Catalog
models are small modifications of these (\citealt{lupton}); details
below.  So our goal here is to provide replacements for these models
in order to improve the performance of galaxy image modeling and
analysis software.  Some of the models we present have been used
previously in tools for precise photometry (\citealt{bundy}), and are
being used in the Large Synoptic Survey Telescope (LSST) analysis
codes along with shapelet representations for the PSF
(\citealt{bosch}).  These models could also be used to speed other
image-fitting systems, like the very successful \project{GALFIT}
(\citealt{galfit}).

@@

In addition to these, there are general S\'ersic (``ser'') profiles, of
which the exp and dev profiles are special cases.  The general ser
profile has one parameter (the ``index'') $n$:
\begin{eqnarray}\displaystyle
Q^{\ser}(\spos\given n) &\equiv& \exp(-\alpha^{\ser}(n)\,[|\spos|^{1/n} - 1])
\\
\{\alpha^{\ser}(2), \alpha^{\ser}(3), \alpha^{\ser}(5)\} &\equiv& \{3.67206075, 5.67016119, 9.66871461\}
\quad ,
\end{eqnarray}
where we have given the constant $\alpha^{\ser}(n)$ for just a few
values of $n$ (\citealt{ciotti}; the $\exp$ and $\dev$ profiles give
the $n=1$ and $n=4$ values).

The \sdss\ pipelines make use of modified profiles, which have been
truncated smoothly at large radius and (in the case of the
de~Vaucouleurs profile) ``softened'' at the center (\citealt{lupton}).
The \sdss\ form of the exponential (lux) profile is
\begin{eqnarray}\displaystyle
Q^{\lux}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha^{\lux}\,[|\spos| - 1]) & \mbox{for~}|\spos| < 3 \\
  \exp(-\alpha^{\lux}\,[|\spos| - 1])
  \,\left[1 - [|\spos| - 3]^2\right]^2 & \mbox{for~}3 < |\spos| < 4 \\
  0                                   & \mbox{for~}4 < |\spos|
\end{array}\right.
\\
\alpha^{\lux} &\equiv& 1.67835
\quad ,
\end{eqnarray}
and the \sdss\ form of the de~Vaucouleurs (luv) profile is
\begin{eqnarray}\displaystyle
Q^{\luv}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha^{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right]) & \mbox{for~}|\spos| < 7 \\
  \exp(-\alpha^{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right])
  \,\left[1 - [|\spos| - 7]^2\right]^2 & \mbox{for~}7 < |\spos| < 8 \\
  0                                   & \mbox{for~}8 < |\spos|
\end{array}\right.
\\
\alpha^{\luv} &\equiv& 7.66925
\quad .
\end{eqnarray}
The half-light inverse-radius parameters $\alpha^{\exp}$ and
$\alpha^{\dev}$---and the softening and cutoff radius parameters---are
taken from the \sdss\ codebase.

@@

The value of these MoG galaxy components comes when they are to be
convolved with a PSF (usually in fact a pixel-convolved PSF) that is
itself also represented as a MoG.  In this scenario, the PSF
$\psi(\Delta\pos)$---which is thought of as a function of focal-plane
displacement $\Delta\pos$ away from, say, a true stellar position---is
represented as a $K$-Gaussian MoG
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: A star of flux $S_s$ at focal-plane position
$\pos_s$ will lead to an image (PSF-convolved intensity map) of the
form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a MoG, any image of a
star---or indeed any image of any set of stars---is also represented
as a MoG.

Applying this PSF to an exp or dev galaxy is slightly more
complicated, because the galaxy has not just a flux $S_g$ and a
central position $\pos_g$; it also has a shape.  Because we are only
considering these simple galaxies, we are only permitting ellipsoidal
shapes, which can be represented by a semi-major axis $a$, a
semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or equivalently by
an affine transformation $\affine_g$ that takes a circle to the
relevant ellipse (and is a therefore a general representation of an
ellipse; it is also the matrix square root of the symmetric variance
tensor describing the ellipse).  The galaxy is distorted by this
affine transformation \emph{prior} to PSF convolution, so the
focal-plane image (PSF-convolved intensity field) for a general (say)
$\exp$ galaxy is given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\affine_g) &=& \sum_{k=1}^K \sum_{m=1}^{M^{\exp}} S_g\,a^{\exp}_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var^{\exp}_m\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Implicitly all vectors are two-dimensional
column vectors, and $\affine_g$ is a $2\times 2$ affine transformation
matrix that contains the ``shape'' (position angle, major-axis, and
ellipticity) information about the galaxy.  The $\dev$, $\luv$, and
$\lux$ cases are all similar.  Note the important and key result of
this \documentname, to wit, that a MoG galaxy model (with $M$
components) convolved with a MoG PSF model (with $K$ components)
yields a MoG model image (with $[M\,K]$ components).  In
\figurename~\ref{fig:example} we show how we are using these MoG
approximations in \project{The Tractor}---a generative modeling
framework for measuring astronomical objects---to render PSF-convolved
galaxy images.

@@

\clearpage
\begin{figure}
\includegraphics[width=\figurewidth]{ser2_K10_MR08_profile_log.pdf}%
\includegraphics[width=\figurewidth]{ser3_K10_MR08_profile_log.pdf}\\
\includegraphics[width=\figurewidth]{ser5_K10_MR08_profile_log.pdf}%
\includegraphics[width=\figurewidth]{mixtures_vs_model_K10.pdf}
\caption{Three ser profiles---with $n=2$, $3$, and $5$---and
  approximations.  The top-left, top-right, and bottom-left panels are
  equivalent to those in the top-right of \figurename~\ref{fig:exp}.
  \textsl{bottom-right:} The dependence on the amplitudes $a^{\ser}_m$
  and root-variances $\sqrt{v^{\ser}_m}$ on ser index $n$.}
\end{figure}

>>>>

% to-do
% -----
% - HOGG: go through remining github issues and close or fix
% - HOGG: send to Bertin and Rix and Peng and Lupton and Bundy for comments
% - HOGG: submit!
% - HOGG: can we drop the regularization in the badness and still succeeed?  If so, let's do that (and drop it in the text).
% - HOGG: make what I say about M=6 lux and M=8 lux true -- ie, edit the Tractor code!
% - DSTN: can we be more specific about how Bosch is using them? I would like to cite users.
% - HOGG: should we make Machine Readable Tables for PASP?  Yes.

@@

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian (MoG)
modeling.  We have pioneered this in the area of distribution modeling
in large numbers of dimensions (\citealt{xd, xdqso, xdqsoz}), where
convolution occurs because the true or noise-free distribution is
convolved with the noise before being observed.  Here we are going to
capitalize on the convolution properties of MoGs in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is always
easier with Gaussians than cuspy profiles, but is \emph{particularly}
useful in data sets in which the PSF itself has also been modeled as a
MoG, which is not uncommon (for example, the \project{Sloan Digital
  Sky Survey} imaging pipelines described by \citealt{lupton} make
approximate MoG PSF models for every imaging field).

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\project{Sloan Digital Sky Survey} (\sdss) Catalog galaxy models,
which only include exponential, de~Vaucouleurs, and composite (mixture
of the two) radial profiles.  In detail, in fact, the \sdss\ Catalog
models are small modifications of these; details
below.  So our goal here is to provide replacements for these models
in order to improve the performance of galaxy image modeling and
analysis software.  Some of the models we present have been used
previously in tools for precise photometry (\citealt{bundy}), and are
being used in the Large Synoptic Survey Telescope (LSST) analysis
codes along with shapelet representations for the PSF
(\citealt{bosch}).  These models could also be used to speed other
image-fitting systems, like the very successful \project{GALFIT}
(\citealt{galfit}).

@@

In addition to these, there are general S\'ersic (``ser'') profiles, of
which the exp and dev profiles are special cases.  The general ser
profile has one parameter (the ``index'') $n$:
\begin{eqnarray}\displaystyle
Q^{\ser}(\spos\given n) &\equiv& \exp(-\alpha^{\ser}(n)\,[|\spos|^{1/n} - 1])
\\
\{\alpha^{\ser}(2), \alpha^{\ser}(3), \alpha^{\ser}(5)\} &\equiv& \{3.67206075, 5.67016119, 9.66871461\}
\quad ,
\end{eqnarray}
where we have given the constant $\alpha^{\ser}(n)$ for just a few
values of $n$ (\citealt{ciotti}; the $\exp$ and $\dev$ profiles given
above provide values for $n=1$ and $n=4$.)

The \sdss\ pipelines (\citealt{lupton}) make use of modified profiles,
which have been truncated smoothly at large radius and (in the case of
the de~Vaucouleurs profile) ``softened'' at the center.  The
\sdss\ form of the exponential (lux) profile is
\begin{eqnarray}\displaystyle
Q^{\lux}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha^{\lux}\,[|\spos| - 1]) & \mbox{for~}|\spos| < 3 \\
  \exp(-\alpha^{\lux}\,[|\spos| - 1])
  \,\left[1 - [|\spos| - 3]^2\right]^2 & \mbox{for~}3 < |\spos| < 4 \\
  0                                   & \mbox{for~}4 < |\spos|
\end{array}\right.
\\
\alpha^{\lux} &\equiv& 1.67835
\quad ,
\end{eqnarray}
and the \sdss\ form of the de~Vaucouleurs (luv) profile is
\begin{eqnarray}\displaystyle
Q^{\luv}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha^{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right]) & \mbox{for~}|\spos| < 7 \\
  \exp(-\alpha^{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right])
  \,\left[1 - [|\spos| - 7]^2\right]^2 & \mbox{for~}7 < |\spos| < 8 \\
  0                                   & \mbox{for~}8 < |\spos|
\end{array}\right.
\\
\alpha^{\luv} &\equiv& 7.66925
\quad .
\end{eqnarray}
The half-light inverse-radius parameters $\alpha^{\exp}$ and
$\alpha^{\dev}$---and the softening and cutoff radius parameters---are
taken from the \sdss\ codebase.

@@

In \figurename~\ref{fig:ser}, we show the dependence of amplitudes
$a^{\ser}_m(n)$ and variances $v^{\ser}_m(n)$ on the ser index $n$.
There is clearly continuity; a valuable follow-up project would be to
give expressions for the amplitudes and variances as a function of ser
index $n$.  In the absence of cleverness our advice would be to make
use of smooth interpolation.

The value of these MoG approximations comes when they are to be
convolved with a PSF (usually in fact a pixel-convolved PSF) that is
itself also represented as a MoG.  In this scenario, the PSF
$\psi(\Delta\pos)$---which is thought of as a function of focal-plane
displacement $\Delta\pos$ away from, say, a true stellar position---is
represented as a $K$-Gaussian MoG
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: A star of flux $S_s$ at focal-plane position
$\pos_s$ will lead to an image (PSF-convolved intensity map) of the
form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a MoG, any image of a
star---or indeed any image of any set of stars---is also represented
as a MoG.

Applying this PSF to an exp or dev galaxy is slightly more
complicated, because the galaxy has not just a flux $S_g$ and a
central position $\pos_g$; it also has a shape.  Because we are only
considering these simple galaxies, we are only permitting ellipsoidal
shapes, which can be represented by a semi-major axis $a$, a
semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or equivalently by
an affine transformation $\affine_g$ that takes a circle to the
relevant ellipse (and is therefore a general representation of an
ellipse; it is also the matrix square root of the symmetric variance
tensor describing the ellipse).  The galaxy is distorted by this
affine transformation \emph{prior} to PSF convolution, so the
focal-plane image (PSF-convolved intensity field) for a general (say)
$\exp$ galaxy is given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\affine_g) &=& \sum_{k=1}^K \sum_{m=1}^{M^{\exp}} S_g\,a^{\exp}_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var^{\exp}_m\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Implicitly all vectors are two-dimensional
column vectors, and $\affine_g$ is a $2\times 2$ affine transformation
matrix that contains the ``shape'' (position angle, major-axis, and
ellipticity) information about the galaxy.  The $\dev$, $\luv$, and
$\lux$ cases are all similar.  Note the important and key result of
this \documentname, to wit, that a MoG galaxy model (with $M$
components) convolved with a MoG PSF model (with $K$ components)
yields a MoG model image (with $[M\,K]$ components).  In
\figurename~\ref{fig:example} we show how we are using these MoG
approximations in \project{The Tractor}---a generative modeling
framework for measuring astronomical objects---to render PSF-convolved
galaxy images.

@@

\clearpage
\begin{figure}
\includegraphics[width=\figurewidth]{ser2_K10_MR08_profile_log.pdf}%
\includegraphics[width=\figurewidth]{ser3_K10_MR08_profile_log.pdf}\\
\includegraphics[width=\figurewidth]{ser5_K10_MR08_profile_log.pdf}%
\includegraphics[width=\figurewidth]{mixtures_vs_model_K10.pdf}
\caption{Three ser profiles---with $n=2$, $3$, and $5$---and
  approximations.  The top-left, top-right, and bottom-left panels are
  equivalent to those in the top-right of \figurename~\ref{fig:exp}.
  \textsl{bottom-right:} The dependence on the amplitudes $a^{\ser}_m$
  and root-variances $\sqrt{v^{\ser}_m}$ on ser index $n$.\label{fig:ser}}
\end{figure}
